Dynamics of Laterally Propagating Flames in X-ray Bursts. II. Realistic Burning \& Rotation

### 2020-12-19 xrb2/paper.tex
Minor edits, added to section 3.3

The mean molecular weight $\bar{A}$ within each of our simulations seems to grow along defined tracks confined to certain temperatures $T$, as shown in the Figure \ref{fig:abar} phase plots. We believe that the tracks in the plot correspond to reaction pathways in the reaction network. Comparing Figure \ref{fig:abar} to \citet{flame_wave1}, these tracks are much more neat and clearly defined. The ''messiness" of the tracks may be dependent on how mixed the flame interior is --- since these new simulations are un-boosted, they may be inherently less mixed than those in \citet{flame_wave1}. {\tt F1000} aligns with this interpretation: its $\bar{A}$ tracks are somewhat disrupted compared to the slower rotation runs, possibly due to the more vigorous mixing of the vortex at the flame front. Comparing the different runs, we also see that as the rotation rate increases, so does the peak temperature. This makes sense if higher rotation leads to a more concentrated, intense vortex near the flame front. It also agrees with our earlier interpretation of enhanced burning seen in Figure \ref{fig:time_series_1000} for {\tt F1000}. The disruption of reaction pathways with higher rotation rates and mixed flames may have interesting consequences; if the reactions are disrupted enough, it is possible that the burning could be snuffed out. This can occur in combustion scenarios when the turbulent vortices become so small they penetrate the reaction zone and completely disrupt the burning.

@@

We found in \citet{flame_wave1} that the flame speed $s$ obeys:
\begin{equation}
	\label{eqn:rossby}
	s \appropto L_R (\epsdot)^{\half} \propto \frac{(\epsdot)^{\half}}{\Omega}
\end{equation}
where $\epsdot$ is the specific energy generation rate and $L_R$ is the Rossby length. This was consistent with the results of \citet{cavecchi:2013}, who noted that the flame speeds in their simulations scaled with the Rossby length. As seen in Figure~\ref{fig:flame_speeds_1}, there is no obvious inverse scaling of the flame speed with rotation rate in this set of runs. We observed earlier, however, that nuclear reactions progress more quickly at the higher rotation rate. This results in a higher $\enucdot$ --- up to four to six times higher near the burning front after the flame ignites --- which may counteract the reduction in flame speed from the increased Coriolis confinement. The initial position of $\tt{F1000}$ does appear to be reduced compared to the slower $\tt{F500}$. From Equation \ref{eqn:rossby}, increasing the rotation rate from 500 Hz to 1000 Hz should decrease $L_R$ by a factor of two, and the initial position of the flame should be reduced by a similar factor. Qualitatively, we think that the higher rotation $\tt{F1000}$ run starts with a tighter votex due to the increased Coriolis force, leading to a lower initial position for the flame front. $L_R$ is not, however, a perfect metric for the flame lengthscale. The scaling described in Equation \ref{eqn:rossby} seems to be more accurate at higher rotation rates (1000 Hz and up), based on \citet{flame_wave1}. We also observe that $\tt{F500}$ accelerates faster than $\tt{F1000}$, which appears to experience a small deceleration at early times. \MarginPar{This is speculation on my part -- does someone else have clear idea why?} This disparity may be a direct result of the difference in Coriolis force. It could also be due to the 500 Hz flame taking more time to develop fully and reach higher burning rates, resulting in an $\epsdot$ \MarginPar{Is this supposed to be enucdot???}increase over time.

@@

To explore the effect of different initial temperature configurations, we run four simulations fixed at a rotation rate of 500 Hz with temperatures as shown in Table \ref{table:runs}. For all the 500 Hz simulations (with the exception of the coolest run, $\tt{F500}$), we set $T_\mathrm{star}$ = $T_\mathrm{hi}$, scaling $\delta T$ accordingly to maintain a consistent value of $T_\mathrm{hi} + \delta T$. If we let $T_\mathrm{star} < T_\mathrm{hi}$, the cooler neutron star surface might act as a heat sink and siphon away energy that would otherwise go into heating the burning layer.

@@

Figure~\ref{fig:compare_500Hz_abar} shows three 500 Hz simulations with different initial temperature structures. The hottest of the three models, $\tt{F500\_3E8}$, has a faster propagating flame. It also appears to have a secondary burning region above the neutron star atmosphere layer (the atmosphere edge is located at $\sim 1.2\times 10^4$ cm). It is likely that a small amount of hot material broke out of the flame front and rose through the atmosphere. This effect, however, may be numerical rather than physical, resulting from the boundary conditions imposed at the axis of symmetry for these simulations.

In contrast to the models with $T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$, $\tt{F500\_4E8}$ is so hot that the organized flame front structure is lost. The fuel layer undergoes stable burning and essentially simmers in place, rather than forming a flame; this effect can occur in nature, for neutron stars with accretion that balances the Eddington luminosity \citep{fujimoto1981,bildsten1998thermonuclear,keek2009effect}. This model burns so strongly that it is only run out to $40$ ms. Due to the lack of a clear burning front, we neglect to analyze the flame propagation speed and acceleration for this run. After an initial period of burning moves across the domain, residual burning continues and eventually ignites the entire fuel layer at late times, as shown in Figures \ref{fig:4e8_stacked_enuc} and \ref{fig:4e8_stacked_abar} for three snapshots taken in the last ten seconds of the simulation. In Figure~\ref{fig:4e8_stacked_enuc}, $\enucdot$ reaches values of $10^{18} - 10^{20}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ across the domain and at heights up to $\sim 0.5\times 10^4$ cm. There is still significant burning occuring even higher, with $\enucdot$ reaching $\sim 10^{15}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ at heights up to $\sim 0.8\times 10^4$ cm. For comparison, the next hottest run ($\tt{F500\_3E8}$) only reaches maximum $\enucdot$ values of $10^{18}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ and a maximum height of $\sim 0.5\times 10^4$ cm, even at the latest timesteps ($\sim 70~\mathrm{ms}$, when the flame is most developed). Significant $\enucdot$ values for $\tt{F500\_3E8}$ are confined to the flame front, rather than spanning the entire domain.
\MarginPar{NF: should I include a plot for this or is it ok to just say it?}. 
Figure~\ref{fig:4e8_stacked_abar} plots $\bar{A}$ for $\tt{F500\_4E8}$, and similarly shows burning that extends across the domain and high into the atmosphere and fuel layer, lacking the charactersitic flame structure shown in $\bar{A}$ plots for the lower temperature $500$ Hz runs (see Figure~\ref{fig:compare_500Hz_abar}). A distinct mass of material - resembling a more dramatic version of that shown in Figure~\ref{fig:compare_500Hz_abar} - appears to have broken out of the atmosphere near the axis of symmetry. Again, this may just be a result of the imposed boundary conditions rather than a physical effect. %Can also add a note here about how high abar values are reached across the domain, maybe don or alice can help give a more detailed analysis of abar. Maybe I could also segue into Alice's phase plots

@@

Looking at the phase plots in Figures~\ref{fig:urho_hot}-\ref{fig:abar_hot}, we see:
\begin{itemize}
    \item The $u$-$\rho$ plot in Figure~\ref{fig:urho_hot} shows that the horizontal velocity when $T_{\rm star} = 3\times10^8$~K is sufficiently greater than for the two cooler runs. This is consistent with the plot of the flame front position, where we saw that the hottest flame was significantly faster at later times than the cooler flames.
    \item The $u$-$v$ plot in Figure~\ref{fig:uv_hot} further demonstrates this, as again we see that the magnitude of the velocity of the flame for the hottest run is significantly larger. Also of interest in these plots is that the coolest run does not seem to have developed the characteristic vortex structure at the burning front (i.e. where $\dot{e}_{\rm nuc}$ is greatest) that can be clearly seen for the hotter runs.
    \item The $\overline{A}$-$T$ plot in Figure~\ref{fig:abar_hot} suggests that the actual reactions happening in the hottest run may be different than for the cooler runs. Not only is the flame itself significantly hotter overall, but there is a large causally connected region across a wide range of $\overline{A}$. This indicates that the hotter temperature has facilitated significant burning in reaction pathways that were not favored at the lower temperatures.
\end{itemize}

@@

The method for measuring flame propagation described in Section \ref{ssec:rot_propagation} is applied in Figure \ref{fig:flame_speeds_2} to the three 500 Hz runs with $T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$. As the initial $T_{\mathrm{star}}$ is increased, the flame becomes greatly accelerated. The initial flame velocities and accelerations derived from a quadratic least-squares fit to each of the datasets are provided in Table \ref{table:flame_speeds}. The reason for the acceleration of the flame over time is not entirely clear. It could be that energy released from burning begins to dominate the flame's propagation as it evolves, increasing the flame speed over time. Another possibility is that the increased temperatures lead to enhanced turbulent mixing effects that pull in more fresh fuel for the flame to burn. Yet another possibility is that the higher initial $T_{\mathrm{star}}$ leads to a greater average temperature in the fuel layer over time, making it easier for the flame to burn the fuel and propagate.

>>>>

The mean molecular weight $\bar{A}$ within each of our simulations seems to grow along defined tracks confined to certain temperatures $T$, as shown in the Figure \ref{fig:abar} phase plots. We believe that the tracks in the plot correspond to reaction pathways in the reaction network. Comparing Figure \ref{fig:abar} to \citet{flame_wave1}, these tracks are much more neat and clearly defined. The ''messiness" of the tracks may be dependent on how mixed the flame interior is --- since these new simulations are un-boosted, they may be inherently less mixed than those in \citet{flame_wave1}. {\tt F1000} aligns with this interpretation: its $\bar{A}$ tracks are somewhat disrupted compared to the slower rotation runs, possibly due to the more vigorous mixing of the vortex at the flame front. Comparing the different runs, we also see that as the rotation rate increases, so does the peak temperature. This makes sense if higher rotation leads to a more concentrated, intense vortex near the flame front. It also agrees with our earlier interpretation of the enhanced burning seen in Figure \ref{fig:time_series_1000} for {\tt F1000}. The disruption of reaction pathways with higher rotation rates and mixed flames may have interesting consequences; if the reactions are disrupted enough, it is possible that the burning could be snuffed out. This can occur in combustion scenarios when the turbulent vortices become so small they penetrate the reaction zone and completely disrupt the burning.

@@

We found in \citet{flame_wave1} that the flame speed $s$ obeys:
\begin{equation}
	\label{eqn:rossby}
	s \appropto L_R (\epsdot)^{\half} \propto \frac{(\epsdot)^{\half}}{\Omega}
\end{equation}
where $\epsdot$ is the specific energy generation rate and $L_R$ is the Rossby length. This was consistent with the results of \citet{cavecchi:2013}, who noted that the flame speeds in their simulations scaled with the Rossby length. As seen in Figure~\ref{fig:flame_speeds_1}, there is no obvious inverse scaling of the flame speed with rotation rate in this set of runs. We observed earlier, however, that nuclear reactions progress more quickly at the higher rotation rate. This results in a higher $\enucdot$ --- up to four to six times higher near the burning front after the flame ignites --- which may counteract the reduction in flame speed from the increased Coriolis confinement. The initial position of $\tt{F1000}$ does appear to be reduced compared to the lower rotation $\tt{F500}$ run. From Equation \ref{eqn:rossby}, increasing the rotation rate from 500 Hz to 1000 Hz should decrease $L_R$ by a factor of two, and the initial position of the flame should be reduced by a similar factor (again, due to greater confinement from the Coriolis force). $L_R$ is not, however, a perfect metric for the flame lengthscale. The scaling described in Equation \ref{eqn:rossby} seems to be more accurate at higher rotation rates (1000 Hz and up), based on \citet{flame_wave1}. We also observe that $\tt{F500}$ accelerates faster than $\tt{F1000}$, which appears to experience a small deceleration at early times. \MarginPar{This is speculation on my part -- does someone else have clear idea why?} This disparity may be a direct result of the difference in Coriolis force. It could also be due to the 500 Hz flame taking more time to develop fully and reach higher burning rates, resulting in an $\epsdot$ \MarginPar{Is this supposed to be enucdot???}increase over time.

@@

 \MarginPar{NF: I think we should plot peak enucdot as a function of time for each run, it would be useful for me to reference it in one of the later sections}

To explore the effect of different initial temperature configurations, we run four simulations fixed at a rotation rate of 500 Hz with temperatures as shown in Table \ref{table:runs}. For all the 500 Hz simulations (with the exception of the coolest run, $\tt{F500}$), we set $T_\mathrm{star}$ = $T_\mathrm{hi}$, scaling $\delta T$ accordingly to maintain a consistent value of $T_\mathrm{hi} + \delta T$. If we let $T_\mathrm{star} < T_\mathrm{hi}$, the cooler neutron star surface might act as a heat sink and siphon away energy that would otherwise go into heating the burning layer. By choosing $T_\mathrm{star}$ = $T_\mathrm{hi}$, we can effectively explore simulations with greater surface heating. There are two physically distinct mechanisms that could each produce an increased temperature at the crust: crustal heating or a higher accretion rate. In these simulations, we cannot distinguish between whether we are generating heat due to a higher accretion rate or due to a shallow crustal heating mechanism, because we are not resolving the crust. Instead, we are modeling the resulting heating effect.

@@

Figure~\ref{fig:compare_500Hz_abar} shows $\bar{A}$ for three 500 Hz simulations with different initial temperature structures at $t = 70~\mathrm{ms}$. The hottest of the three models, $\tt{F500\_3E8}$, has a faster propagating flame (discussed further in Section \ref{ssec:temp_prop}). It also appears to have a secondary burning region (colored light orange) forming on the far left above the flame edge. It is likely that a small amount of hot material broke out of the flame and rose through the atmosphere. This effect, however, may be numerical rather than physical, resulting from the boundary conditions imposed at the axis of symmetry for these simulations.

Similar to the $\bar{A}$ trends found for different rotation rates in Figure \ref{fig:abar}, the $\bar{A}$-$T$ phase plot in Figure~\ref{fig:abar_hot} suggests that the actual reactions happening in the hottest run may be different than for the cooler runs. Not only is the $\tt{F500\_3E8}$ flame itself significantly hotter overall, but there is a large causally connected region across a wide range of $\bar{A}$. This indicates that the hotter temperature has facilitated significant burning in reaction pathways that were not favored at the lower temperatures. The reaction pathways are also very disrupted for $\tt{F500\_3E8}$ compared to the cooler runs, suggesting that the hotter temperature leads to more turbulent mixing. Compared to the highest rotation rate run, $\tt{F1000}$, $\tt{F500\_3E8}$ reaches a peak temperature roughly $1 \times 10^9~\mathrm{K}$ higher ($\sim 2.7 \times 10^9~\mathrm{K}$ versus $\sim 1.7 \times 10^9~\mathrm{K}$). Though increasing the rotation rate does seem to concentrate the burning and increase the peak temperature, it does not reach the same level as directly increasing the temperature.

In contrast to the models with $T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$, $\tt{F500\_4E8}$ is so hot that the organized flame structure is lost. The fuel layer undergoes stable burning and essentially simmers in place, rather than forming a flame; this effect can occur in nature, for neutron stars with accretion that balances the Eddington luminosity \citep{fujimoto1981,bildsten1998thermonuclear,keek2009effect}. This model burns so strongly that it is only run out to $40$ ms. After an initial period of burning moves across the domain, residual burning continues and eventually ignites the entire fuel layer at late times, as shown in Figures \ref{fig:4e8_stacked_enuc} and \ref{fig:4e8_stacked_abar} for three snapshots taken in the last ten seconds of the simulation. In Figure~\ref{fig:4e8_stacked_enuc}, $\enucdot$ reaches values of $10^{18} - 10^{20}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ across the domain and at heights up to $\sim 0.5\times 10^4$ cm. There is still significant burning occuring even higher, with $\enucdot$ reaching $\sim 10^{15}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ at heights up to $\sim 0.8\times 10^4$ cm. For comparison, the next hottest run ($\tt{F500\_3E8}$) only reaches maximum $\enucdot$ values of $10^{18}~\mathrm{erg}\,\mathrm{g}^{-1} \mathrm{s}^{-1}$ and a maximum height of $\sim 0.5\times 10^4$ cm, even at the latest timesteps ($\sim 70~\mathrm{ms}$, when the flame is most developed). Significant $\enucdot$ values for $\tt{F500\_3E8}$ are confined to the flame front, rather than spanning the entire domain.
\MarginPar{NF: should I include a plot for this or is it ok to just say it?}. 
Figure~\ref{fig:4e8_stacked_abar} shows $\bar{A}$ for $\tt{F500\_4E8}$. Again, burning extends across the domain and high into the atmosphere and fuel layer, lacking the charactersitic flame structure shown in $\bar{A}$ plots for the lower temperature $500$ Hz runs (see Figure~\ref{fig:compare_500Hz_abar}). A distinct mass of material - resembling a more dramatic version of that shown in Figure~\ref{fig:compare_500Hz_abar} - appears to have broken out of the atmosphere near the axis of symmetry. The atmosphere edge is located at $\sim 1.2\times 10^4~\mathrm{cm}$. To reiterate, this may just be a result of the imposed boundary conditions rather than a physical effect. $\tt{F500\_4E8}$ clearly reaches much higher $\bar{A}$ values across the domain (especially at the latest snapshot, $t = 40~\mathrm{ms}$) compared to all the other runs described in this paper (see Figures \ref{fig:compare_500Hz_abar}, \ref{fig:time_series_500}, and \ref{fig:time_series_1000}). This is not surprising, since as we have seen through our varied temperature simulations, the burning is very temperature-sensitive.

@@

The method for measuring flame propagation described in Section \ref{ssec:rot_propagation} is applied in Figure \ref{fig:flame_speeds_2} to the three 500 Hz runs with $T_{\mathrm{star}} \leq 3 \times 10^8~\mathrm{K}$. Due to the lack of a clear burning front in $\tt{F500\_4E8}$, we neglect to analyze its propagation speed and acceleration. As the initial $T_{\mathrm{star}}$ is increased, the flame becomes greatly accelerated. The initial flame velocities and accelerations derived from a quadratic least-squares fit to each of the datasets are provided in Table \ref{table:flame_speeds}. The reason for the acceleration of the flame over time is not entirely clear. It could be that energy released from burning begins to dominate the flame's propagation as it evolves, increasing the flame speed over time. Another possibility is that the increased temperatures lead to enhanced turbulent mixing effects that pull in more fresh fuel for the flame to burn. Yet another possibility is that the higher initial $T_{\mathrm{star}}$ leads to a greater average temperature in the fuel layer over time, making it easier for the flame to burn the fuel and propagate.

Looking at the phase plots in Figures~\ref{fig:urho_hot}-\ref{fig:abar_hot}, we see:
\begin{itemize}
    \item The $u$-$\rho$ plot in Figure~\ref{fig:urho_hot} shows that the horizontal velocity when $T_{\rm star} = 3\times10^8$~K is sufficiently greater than for the two cooler runs. This is consistent with the plot of the flame front position, where we saw that the hottest flame was significantly faster at later times than the cooler flames.
    \item The $u$-$v$ plot in Figure~\ref{fig:uv_hot} further demonstrates this, as again we see that the magnitude of the velocity of the flame for the hottest run is significantly larger. Also of interest in these plots is that the coolest run does not seem to have developed the characteristic vortex structure at the burning front (i.e. where $\dot{e}_{\rm nuc}$ is greatest) that can be clearly seen for the hotter runs.
\end{itemize}

@@
